Distributional Trees for Circular Data

### 2019-01-30 pkg/disttree/inst/IWSM2019/abstract_IWSM2019.tex
Include application part (w/o fig)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2414 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 7f9ff234d28d19800e05dfcff65ac1d78f2f7411

\textit{TO DO}

>>>>

%\begin{figure}[!ht]\centering
%\includegraphics[width = 1.\textwidth]{circtree_plot_small.pdf}
%\caption{\label{fig:circtree} Fitted distribution tree for a circular response employing the von Mises distribution.}\label{fig:tree}
%\end{figure}

Wind is a classical circular quantity and accurate forecasts of wind direction
are of great importance for decision making processes and risk management,
e.g., in air traffic management and renewable energy production. As a suitable
and useful application of circular regression trees, we therefore forecast the
wind direction at the airport of Innsbruck. As potential split variables we
employ several 1-hourly lagged observed weather variables (e.g., wind speed,
relative humidity, pressure) and derived spatial and temporal difference of
these lagged observations.

%Innsbruck lies with the west-east oriented Inn Valley with an approx. 2300\,m
%high mountain ridge on the northern side and with the south-north oriented Wipp
%Valley between the approx. 2300\,m high mountains on the southern side.
%Therefore, skill full covariates are pressure differences along the Inn Valley
%as indices for up- or down-valley winds. For the typical dry gap wind coming
%from the south, indices are, e.g. low relative humidity or temperature
%differences between Innsbruck and stations up the Wipp Valley.

The weather station of interest, Innsbruck Airport, lies at the bottom of the
Inn Valley, a deep mountain valley in the Eastern European Alps orientated
parallel to the Alpine crest. At Innsbruck, the Inn Valley is bordered in the
North with a approx. 2300\,m high mountain ridge and opens in the South into
the crest-perpendicular Wipp Valley. Therefore, mainly three different wind
regimes exist: Up-valley winds blowing from the valley mouth towards the upper
valley (from east to west), down-valley winds blowing in the direction of the
valley mouth (from west to east), and dry and warm downslope winds blowing
across the Alpine crest along the Wipp Valley towards Innsbruck (from
south-east to north-west).

As an illustration of tree based forecasts of wind direction for Innsbruck Airport, a
model based regression tree of depth two is shown in Fig.~\ref{fig:tree}.
Winds typically blow from areas of high pressure to areas of low pressure. The
first chosen split variable is therefore the pressure difference between
Innsbruck and a station located near the valley mouth. In the left branch of
the tree, with lower pressure in Innsbruck than at the valley mouth, the second
split is performed on the pressure difference between Innsbruck a station
further up the valley. This leads to the final \emph{Node~3} with low pressure
gradients within the Inn Valley and, therefore, no clear partition in the wind
direction. The final \emph{Node~4} contains measurements with a strong pressure
gradient towards the upper Inn Valley leading to up-valley winds from east to
west. The right branch of the tree is split in relative humidity as an index
for the dry downslope winds. Therefore, for a low relative humidity winds from
south-east to north-west are predominant in the final \emph{Node~6}. Whereas,
the final \emph{Node~7} contains the cases with down-valley winds towards the
valley mouth in the East.
